Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-02-20 newpaper/paper.tex
wip

\usepackage{xcolor}
\usepackage{mathpartir}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{scalerel}
\usepackage{bm}
\usepackage{mathtools}
%% \usepackage{amssymb}

\newcommand{\mit}[1]{\mathit{#1}}
\newcommand{\msf}[1]{\mathsf{#1}}
\newcommand{\mbf}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\mdo}{\mbf{do}\,}
\newcommand{\ind}{\hspace{1em}}
\newcommand{\bif}{\mbf{if}\,}
\newcommand{\bthen}{\mbf{then}\,}
\newcommand{\belse}{\mbf{else}\,}
\newcommand{\return}{\mbf{return}\,}
\newcommand{\lam}{\lambda\,}
\newcommand{\data}{\mbf{data}\,}
\newcommand{\where}{\mbf{where}}
\newcommand{\M}{\msf{M}}
\newcommand{\letrec}{\mbf{letrec}\,}
\newcommand{\of}{\mbf{of}\,}
\newcommand{\go}{\mit{go}}
\newcommand{\add}{\mit{add}}
\newcommand{\letdef}{\mbf{let\,}}
\newcommand{\map}{\mit{map}}
\newcommand{\vas}{\mit{as}}
\newcommand{\vbs}{\mit{bs}}
\newcommand{\vxs}{\mit{xs}}
\newcommand{\vys}{\mit{ys}}
\newcommand{\vma}{\mit{ma}}
\newcommand{\vmb}{\mit{mb}}

@@

\newcommand{\Int}{\msf{Int}}
\newcommand{\List}{\msf{List}\,}
\newcommand{\Nil}{\msf{Nil}}
\newcommand{\Cons}{\msf{Cons}}
\newcommand{\Reader}{\msf{Reader}}
\newcommand{\ReaderT}{\msf{ReaderT}}
\newcommand{\Monad}{\msf{Monad}}
\newcommand{\Applicative}{\msf{Monad}}
\newcommand{\class}{\msf{class}}
\newcommand{\Functor}{\msf{Functor}}
\newcommand{\Bool}{\msf{Bool}}
\newcommand{\Statel}{\msf{State}}
\newcommand{\fro}{\leftarrow}
\newcommand{\case}{\mbf{case\,}}

@@

\theoremstyle{remark}
\newtheorem{notation}{Notation}
\newcommand{\id}{\mit{id}}
\newcommand{\mup}{\mit{up}}
\newcommand{\mdown}{\mit{down}}
\newcommand{\tyclass}{\mbf{class}}
\newcommand{\instance}{\mbf{instance}\,}
\newcommand{\Improve}{\msf{Improve}}
\newcommand{\Gen}{\msf{Gen}}
\newcommand{\unGen}{\mit{unGen}}
\renewcommand{\Vec}{\msf{Vec}}
\newcommand{\gen}{\mit{gen}}
\newcommand{\genRec}{\mit{genRec}}
\newcommand{\fmap}{<\!\!\$\!\!>}
\newcommand{\ap}{<\!\!*\!\!>}
\newcommand{\runGen}{\mit{runGen}}
\newcommand{\qt}[1]{\ql#1\qr}
\newcommand{\lift}{\mit{lift}}
\newcommand{\liftGen}{\mit{liftGen}}
\newcommand{\MonadGen}{\msf{MonadGen}}
\newcommand{\RA}{\Rightarrow}
\newcommand{\EitherT}{\msf{EitherT}}
\newcommand{\StateT}{\msf{StateT}}
\newcommand{\Identity}{\msf{Identity}}
\newcommand{\runIdentity}{\mit{runIdentity}}
\newcommand{\newtype}{\mbf{newtype}\,}
\newcommand{\runMaybeT}{\mit{runMaybeT}}
\newcommand{\dlr}{\,\$\,}
\newcommand{\ImproveF}{\msf{ImproveF}}
\newcommand{\ExceptT}{\msf{ExceptT}}

@@

\begin{abstract}
There are many abstraction tools in modern functional programming which heavily
rely on general-purpose compiler optimization to achieve adequate
performance. For example, monadic binding is a higher-order function which
yields runtime closures in the absence of sufficient compile-time inlining and
beta-reductions, thereby significantly degrading performance. In current systems
such as the Glasgow Haskell Compiler, there is no strong guarantee that
general-purpose optimization can eliminate abstraction overheads, and users only
have indirect and fragile control over code generation through inlining
directives and compiler options. In this paper we propose using a two-stage
language to simultaneously get strong code generation guarantees and strong
abstraction features. The object language is a simply typed first-order language
where all function calls are statically known, and which can be compiled without
runtime closures. The compile-time language is a dependent type theory. The two
are integrated in a two-level type theory. We develop some abstraction
tools in this setting. First, we develop monads and monad transformers. Second,
we develop fusion for push and pull streams. Most of our results are also
adapted to a proof-of-concept library in typed Template Haskell.
\end{abstract}

@@

Modern functional programming supports many convenient abstractions. These often
come with significant runtime overheads. Sometimes the overheads are acceptable,
but in other cases compiler optimization is crucial. Monads in Haskell is an
example for the latter. Even the $\Reader$ monad, which is one of the simplest
in terms of implementation, yields large overheads when compiled without
optimizations. Consider the following snippet.
\begin{alignat*}{3}
  &f :: \Int \to \Reader\,\Bool\,\Int \\
  &f\,x = \mdo\{b \fro \msf{ask}; \bif x\, \bthen \return (x + 10)\, \belse \return (x + 20)\}
\end{alignat*}
With optimizations enabled, GHC compiles this roughly to the following:
\begin{alignat*}{3}
  &f :: \Int \to \Bool \to \Int \\
  &f = \lam x\,b.\, \bif b\, \bthen x + 10\, \belse x + 20
\end{alignat*}
Without optimizations we roughly get:
\begin{alignat*}{3}
  &f = \lam x.\,(\bind)\,\msf{MonadReaderDict}\,\msf{ask}\,(\lam b.\,\bif b\\
  &\ind \bthen\return\,\msf{MonadReaderDict}\,(x + 10)\\
  &\ind \belse\hspace{0.25em}\return\,\msf{MonadReaderDict}\,(x + 20))
\end{alignat*}
Here, $\msf{MonadReaderDict}$ is a runtime dictionary, containing the methods of
the $\Monad$ instance for $\Reader$, and $(\bind)\,\msf{MonadReaderDict}$ is a
field projection. Here, a runtime closure will be created for the $\lam b.\,...$
function, and $(\bind)$, $\msf{ask}$ and $\return$ will create additional dynamic
closures.

The difference between optimized and unoptimized code is already large here, and
it gets even larger when we consider monad transformers or code that is
polymorphic over monads. In Haskell, such code is pervasive, even in fairly
basic programs which do not use fancy abstractions. The $\msf{mapM}$ function in
the Haskell Prelude, which maps over a list in some monad, is a third-order and
second-rank polymorphic function in disguise, because its monad dictionary
argument contains the polymorphic second-order $(\bind)$ method.
\begin{alignat*}{3}
  & \msf{mapM} :: \Monad\,m \Rightarrow (a \to m\,b) \to [a] \to m\,[b]
\end{alignat*}
Compiling $\msf{mapM}$ efficiently relies on inlining the instance dictionary,
then inlining the methods contained there, and also inlining the functions
that the higher-order binding method is applied to.

GHC's optimization efforts are respectable, and it has gotten quite good over
its long history of development. However, there is no strong guarantee that
certain optimizations will happen. Control over optimizations remains tricky,
fragile and non-compositional. \texttt{INLINE} and \texttt{REWRITE} pragmas can
be used to control code generation, but without any strong guarantee, and their
sophisticated usage requires knowledge of GHC internals. For example, correctly
specifying the \emph{ordering} of certain rule applications is often needed. We
have to also care about function arities. Infamously, the function composition
operator is defined as $(.)\,f\,g = \lam x \to f\,(g\,x)$ in the base libraries,
instead of as $(.)\,f\,g\,x = f\,(g\,x)$, to get better inlining behavior ---
as explained in a source comment right next to the definition \cite{TODO}.

Although there are numerous tricks and idioms that are used in high-performance
Haskell programming, for reliable performance it is necessary that programmers
periodically \emph{review} GHC's optimized code output. Needless to say, this is
rather time-consuming and poorly scalable.

\subsection{Staged Compilation}

@@

Metaprograms can be deterministic, transparent, and can be run efficiently,
using fast interpreters or machine code compilation. In contrast,
general-purpose optimizers are slower to run, less transparent and less
reliable. Also, metaprogramming allows library authors to exploit
domain-specific optimizations, while it is not realistic for general-purpose
optimizers to know about all domains.

On the other hand, metaprogramming requires some additional care and input from
programmers. Historically, there have been problems with ergonomics as well:
\begin{itemize}
\item Code generation might fail \emph{too late} in the pipeline, producing
      incomprehensible errors; this is often caused by not having enough static
      guarantees about the well-formedness of code output.
\item Tooling for the object language (debugging, profiling, IDE support) often does not
      work for metaprogramming. This is more likely if the object and meta layers are wildly
      different.
\item Metaprogramming may introduce heavy noise and boilerplate, obscuring the logic of
      programs, or impose restrictions on how code can be structured. For instance,
      Template Haskell mandates that metaprograms used in splices in some module are
      defined in a different module.
\end{itemize}

@@

We develop a particular two-level type theory for this purpose, which we
call \textbf{CFTT}, short for ``closure-free type theory''. This consists
of
\begin{itemize}
\item A simply-typed object theory with first-order functions, general recursion and
      finitary algebraic data types. This language is easy to optimize and compile
      downstream in the pipeline, but it lacks many convenience features.
\item A standard Martin-Löf type theory for the compile-time language. This
      allows us to recover many features by metaprogramming.
\end{itemize}

In particular, since the object language is first-order, we guarantee that all
programs in CFTT can be ultimately compiled without any dynamic closures, using
only calls and jumps to statically known code locations. Why focus on closures?
They are the foundation to almost all abstraction tools in functional
programming:
\begin{itemize}
\item Higher-order functions in essentially all functional languages are implemented with closures.
\item Type classes in Haskell use dictionary-passing, which relies on closures for function methods.
\item Functors and first-class modules in OCaml and other ML-s rely on closures.
\end{itemize}
Hence, doing functional programming without closures is a clear demonstration
that we can get rid of abstraction overheads.

It turns out that surprisingly little practical programming relies essentially
on closures. Most of the time, programmers use higher-order functions for
\emph{abstraction}, such as when mapping over lists, where it is expected that
the mapping function will be inlined. In other cases, the use of closures can be
eliminated by small-scale defunctionalization. For example, difference lists
\cite{TODO} are often implemented with closures, but they can be also
implemented as binary trees, with the same programming interface.

\todo{Essential closures: CPS? Threaded interpreters?}

Whole-program defunctionalization is also possible, and it is notably used by
the MLton compiler \cite{TODO}. While this can be practical and effective, it
can be also expensive and require whole-program processing. Also, conceptually
speaking, it does not eliminate closures but instead makes them more transparent
to optimizations. In this paper we do not use defunctionalization.

We note though that our setup is compatible with closures as well, and it can
support two separate type formers for closure-based and non-closure-based
(``static'') functions. Having both of these would be desirable in a practical
system. In the current work we focus on the closure-free case because it is much
less known and developed, and it is quite interesting to see how far we
can go with it.

@@

\todo{CONTRIBUTIONS}

\section{Basics of CFTT}\label{sec:basics-of-cftt}

In the following we give an overview of CFTT features. Here we focus on examples
and informal explanations. We defer the more formal details to Sections
\ref{TODO} and \ref{TODO}. We first review the meta-level language, then the
object-level one, and finally the staging operations which bridge between the
two.

@@

Formally, $\MTy$ is additionally indexed by universes levels, and we have
$\MTy_i : \MTy_{i+1}$. However, this adds a bit of noise, and it is not very
relevant to the current paper, so we shall omit levels. Note that universe
levels have nothing to do with staged compilation; they are about sizing for the
purpose of logical consistency.

We use Agda-like syntax for functions and implicit arguments. A basic example:
\begin{alignat*}{3}
  &\id : \{A : \MTy\} \to A \to A\\
  &\id = \lam x.\, x
\end{alignat*}
Here, the type argument is implicit, and it gets inferred when we use the
function. For example, $id\,\True$ is elaborated to $id\,\{\Bool\}\,\True$,
where the braces mark an explicit application for the implicit argument.
Similarly, we can introduce implicit lambdas explicitly:
\begin{alignat*}{3}
  &id = \lam \{A : \MTy\}(x : A).\,x
\end{alignat*}
We write $(a : A) \times B$ for $\Sigma$-types in $\MTy$. For the field
projections, for $t : (a : A) \times B$, we have $\fst\,t : A$ and $\snd\,t :
B[a \mapsto \fst\,t]$, and pairing is written simply as $(t,\,u)$. Additionally,
we use syntactic sugar for field projections: we can write a type as $(field_1 :
A) \times (field_2 : B) \times (field_3 : C)$, and write $t.field_2$ for a named
field projection. The type itself denotes a right-nested $\Sigma$-type. We can
also write $(t,\,u,\,v)$ for a right-nested iterated pairing. In Haskell style,
we write $()$ for the unit type, and also for its inhabitant.

Inductive types can be introduced using a Haskell-like ADT notation, or with a GADT-style one:
\begin{alignat*}{3}
  &                                           &&\hspace{4em}\data\,\Bool_\M &&: \MTy\,\where\\
  & \data\,\Bool_\M : \MTy = \True\,|\,\False &&\hspace{4em}\ind\ind \True &&: \Bool_\M\\
  &                                           &&\hspace{4em}\ind\ind \False &&: \Bool_\M
\end{alignat*}
Note that we added an $_\M$ subscript to the type; when analogous types can be
defined both on the meta and object levels, we will sometimes use this subscript
to disambiguate the meta-level version. We will also use Haskell-like newtype
notation, such as in $\newtype \msf{Wrap}\,A = \msf{Wrap}\,\{\mit{unWrap} : A
\}$.

Importantly, we can only produce meta-level values by induction on meta-level
data. In general, all construction and elimination rules for type formers in
$\MTy$ stay within $\MTy$.

@@

Since $\VTy$ is a sub-universe of $\Ty$, we have that when $A : \VTy$ then also
$A : \Ty$. Formally, this is specified as an explicit embedding operation but
informally it is nicer to have an implicit subtyping.

>>>>

\usepackage{xcolor}
\usepackage{mathpartir}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{scalerel}
\usepackage{bm}
\usepackage{mathtools}

\newcommand{\mit}[1]{{\mathsf{#1}}}
\newcommand{\msf}[1]{{\mathsf{#1}}}
\newcommand{\mbf}[1]{{\mathbf{#1}}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\mdo}{\mbf{do}\,}
\newcommand{\ind}{\hspace{1em}}
\newcommand{\bif}{\mbf{if}\,}
\newcommand{\bthen}{\mbf{then}\,}
\newcommand{\belse}{\mbf{else}\,}
\newcommand{\return}{\mbf{return}\,}
\newcommand{\pure}{\mbf{pure}\,}
\newcommand{\lam}{\lambda\,}
\newcommand{\data}{\mbf{data}\,}
\newcommand{\where}{\mbf{where}}
\newcommand{\M}{\msf{M}}
\newcommand{\letrec}{\mbf{letrec}\,}
\newcommand{\of}{\mbf{of}\,}
\newcommand{\go}{\mit{go}}
\newcommand{\add}{\mit{add}}
\newcommand{\letdef}{\mbf{let\,}}
\newcommand{\map}{\mit{map}}

\newcommand{\vas}{\mathsf{as}}
\newcommand{\vbs}{\mathsf{bs}}
\newcommand{\vcs}{\mathsf{cs}}
\newcommand{\vxs}{\mathsf{xs}}
\newcommand{\vys}{\mathsf{ys}}
\newcommand{\vma}{\mathsf{ma}}
\newcommand{\vm}{\mathsf{m}}
\newcommand{\vn}{\mathsf{n}}
\newcommand{\vk}{\mathsf{k}}
\newcommand{\vA}{\mathsf{A}}
\newcommand{\vB}{\mathsf{B}}
\newcommand{\vC}{\mathsf{C}}
\newcommand{\vS}{\mathsf{S}}
\newcommand{\vF}{\mathsf{F}}
\newcommand{\vR}{\mathsf{R}}
\newcommand{\vM}{\mathsf{M}}
\newcommand{\vmb}{\mathsf{mb}}
\newcommand{\mAs}{\mathsf{As}}
\newcommand{\va}{\mathsf{a}}
\newcommand{\vb}{\mathsf{b}}
\newcommand{\vc}{\mathsf{c}}
\newcommand{\vd}{\mathsf{d}}
\newcommand{\vx}{\mathsf{x}}
\newcommand{\vy}{\mathsf{y}}
\newcommand{\vz}{\mathsf{z}}
\newcommand{\vf}{\mathsf{f}}
\newcommand{\vfs}{\mathsf{fs}}
\newcommand{\vg}{\mathsf{g}}
\newcommand{\vh}{\mathsf{h}}
\newcommand{\vt}{\mathsf{t}}
\newcommand{\vs}{\mathsf{s}}
\newcommand{\vr}{\mathsf{r}}
\newcommand{\vu}{\mathsf{u}}

\newcommand{\SOP}{\msf{SOP}}
\newcommand{\El}{\msf{El}}
\newcommand{\USOP}{\msf{U}_{\msf{SOP}}}
\newcommand{\Uprod}{\msf{U_P}}
\newcommand{\Elprod}{\msf{El_{P}}}
\newcommand{\IsSOP}{\msf{IsSOP}}
\newcommand{\forEach}{\msf{forEach}}
\newcommand{\single}{\msf{single}}
\newcommand{\msplit}{\msf{split}}
\newcommand{\mapGen}{\msf{mapGen}}

@@

\newcommand{\Int}{\msf{Int}}
\newcommand{\List}{\msf{List}}
\newcommand{\Nil}{\msf{Nil}}
\newcommand{\Cons}{\msf{Cons}}
\newcommand{\Reader}{\msf{Reader}}
\newcommand{\ReaderT}{\msf{ReaderT}}
\newcommand{\Monad}{\msf{Monad}}
\newcommand{\Applicative}{\msf{Applicative}}
\newcommand{\class}{\msf{class}}
\newcommand{\Functor}{\msf{Functor}}
\newcommand{\Bool}{\msf{Bool}}
\newcommand{\Statel}{\msf{State}}
\newcommand{\fro}{\leftarrow}
\newcommand{\case}{\mbf{case\,}}
\newcommand{\foldr}{\msf{foldr}}
\newcommand{\foldl}{\msf{foldl}}
\newcommand{\rep}{\msf{rep}}
\newcommand{\concatMap}{\msf{concatMap}}

@@

\theoremstyle{remark}
\newtheorem{notation}{Notation}
\newtheorem*{axiom}{Axiom}

\newcommand{\id}{\mit{id}}
\newcommand{\mup}{\mit{up}}
\newcommand{\mdown}{\mit{down}}
\newcommand{\tyclass}{\mbf{class}}
\newcommand{\instance}{\mbf{instance}\,}
\newcommand{\Improve}{\msf{Improve}}
\newcommand{\Gen}{\msf{Gen}}
\newcommand{\unGen}{\mit{unGen}}
\renewcommand{\Vec}{\msf{Vec}}
\newcommand{\gen}{\mit{gen}}
\newcommand{\genRec}{\mit{genRec}}
\newcommand{\fmap}{<\!\!\$\!\!>}
\newcommand{\ap}{{<\!\!*\!\!>}}
\newcommand{\runGen}{\mit{runGen}}
\newcommand{\qt}[1]{\ql#1\qr}
\newcommand{\lift}{\mit{lift}}
\newcommand{\liftGen}{\mit{liftGen}}
\newcommand{\MonadGen}{\msf{MonadGen}}
\newcommand{\MonadState}{\msf{MonadState}}
\newcommand{\MonadReader}{\msf{MonadReader}}
\newcommand{\RA}{\Rightarrow}
\newcommand{\EitherT}{\msf{EitherT}}
\newcommand{\Either}{\msf{Either}}
\newcommand{\Left}{\msf{Left}}
\newcommand{\Right}{\msf{Right}}
\newcommand{\StateT}{\msf{StateT}}
\newcommand{\Identity}{\msf{Identity}}

\newcommand{\Stop}{\msf{Stop}}
\newcommand{\Skip}{\msf{Skip}}
\newcommand{\Yield}{\msf{Yield}}

\newcommand{\runIdentity}{\mit{runIdentity}}
\newcommand{\runReaderT}{\mit{runReaderT}}
\newcommand{\newtype}{\mbf{newtype}\,}
\newcommand{\runMaybeT}{\mit{runMaybeT}}
\newcommand{\runStateT}{\mit{runStateT}}
\newcommand{\runState}{\mit{runState}}
\newcommand{\dlr}{\,\$\,}
\newcommand{\ImproveF}{\msf{ImproveF}}
\newcommand{\ExceptT}{\msf{ExceptT}}
\newcommand{\State}{\msf{State}}
\newcommand{\SumVS}{\msf{SumVS}}
\newcommand{\ProdCS}{\msf{ProdCS}}
\newcommand{\Here}{\msf{Here}}
\newcommand{\There}{\msf{There}}
\newcommand{\IsSumVS}{\msf{IsSumVS}}
\newcommand{\MonadJoin}{\msf{MonadJoin}}
\newcommand{\Stream}{\msf{Stream}}
\newcommand{\join}{\mit{join}}
\newcommand{\modify}{\mit{modify}}
\newcommand{\get}{\mit{get}}
\newcommand{\mput}{\mit{put}}
\newcommand{\Rep}{\mit{Rep}}
\newcommand{\encode}{\mit{encode}}
\newcommand{\decode}{\mit{decode}}
\newcommand{\mindex}{\mit{index}}
\newcommand{\mtabulate}{\mit{tabulate}}
\newcommand{\States}{\mit{States}}
\newcommand{\seed}{\mit{seed}}
\newcommand{\step}{\mit{step}}
\newcommand{\Step}{\msf{Step}}
\newcommand{\Pull}{\msf{Pull}}
\newcommand{\MkPull}{\msf{MkPull}}

@@

\begin{abstract}
Many abstraction tools in functional programming rely heavily on general-purpose
compiler optimization to achieve adequate performance. For example, monadic
binding is a higher-order function which yields runtime closures in the absence
of sufficient compile-time inlining and beta-reductions, thereby significantly
degrading performance. In current systems such as the Glasgow Haskell Compiler,
there is no strong guarantee that general-purpose optimization can eliminate
abstraction overheads, and users only have indirect and fragile control over
code generation through inlining directives and compiler options. We propose a
two-stage language to simultaneously get strong guarantees about code generation
and strong abstraction features. The object language is a simply-typed
first-order language which can be compiled without runtime closures. The
compile-time language is a dependent type theory. The two are integrated in a
two-level type theory.

We demonstrate two applications of the system. First, we develop monads and
monad transformers. Here, abstraction overheads are eliminated by staging and we
can reuse almost all definitions from the existing Haskell ecosystem. Second,
we develop pull-based stream fusion. Here we make essential use of dependent
types to give a concise definition of a $\mathsf{concatMap}$ operation with
guaranteed fusion. We provide an Agda implementation and a typed Template
Haskell implementation of these developments.

\end{abstract}

@@

Modern functional programming supports many convenient abstractions. These often
come with significant runtime overheads. Sometimes the overheads are acceptable,
but in other cases compiler optimization is crucial. Monads in Haskell is an
example for the latter. Even the $\Reader$ monad, which is one of the simplest
in terms of implementation, yields large overheads when compiled without
optimizations. Consider the following:
\begin{alignat*}{3}
  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vx\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
\end{alignat*}
With optimizations enabled, GHC compiles this roughly to the code below:
\begin{alignat*}{3}
  &\vf :: \Int \to \Bool \to \Int \\
  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
\end{alignat*}
Without optimizations we roughly get:
\begin{alignat*}{3}
  &\vf = \lam \vx.\,(\bind)\,\msf{MonadReaderDict}\,\msf{ask}\,(\lam \vb.\,\bif \vb\\
  &\ind \bthen\return\,\msf{MonadReaderDict}\,(\vx + 10)\\
  &\ind \belse\hspace{0.25em}\return\,\msf{MonadReaderDict}\,(\vx + 20))
\end{alignat*}
Here, $\msf{MonadReaderDict}$ is a runtime dictionary, containing the methods of
the $\Monad$ instance for $\Reader$, and $(\bind)\,\msf{MonadReaderDict}$ is a
field projection. Here, a runtime closure will be created for the $\lam \vb.\,...$
function, and $(\bind)$, $\msf{ask}$ and $\return$ will create additional dynamic
closures.

The difference between optimized and unoptimized code is already large here, and
it gets even larger when we consider monad transformers or code that is
polymorphic over monads. In Haskell, such code is pervasive, even in fairly
basic programs which do not use fancy abstractions. Consider the $\msf{mapM}$
function from the Haskell Prelude:
\begin{alignat*}{3}
  & \msf{mapM} :: \Monad\,\vm \Rightarrow (\va \to \vm\,\vb) \to [\va] \to \vm\,[\vb]
\end{alignat*}
This is a third-order and second-rank polymorphic function in disguise, because
its monad dictionary argument contains the polymorphic second-order $(\bind)$
method. Compiling $\msf{mapM}$ efficiently relies on inlining the instance
dictionary, then inlining the methods contained there, and also inlining the
functions that the higher-order binding method is applied to.

GHC's optimization efforts are respectable, and it has gotten quite good over
its long history of development. However, there is no strong guarantee that
certain optimizations will happen. Control over optimizations remains tricky,
fragile and non-compositional. \texttt{INLINE} and \texttt{REWRITE} pragmas can
be used to control code generation, but without any strong guarantee, and their
advanced usage requires knowledge of GHC internals. For example, correctly
specifying the \emph{ordering} of certain rule applications is often needed. We
also have to care about formal function arities. Infamously, the function
composition operator is defined as $(.)\,\vf\,\vg = \lam \vx \to \vf\,(\vg\,\vx)$ in the
base libraries, instead of as $(.)\,\vf\,\vg\,\vx = \vf\,(\vg\,\vx)$, to get better inlining
behavior \cite{TODO}. It remains a common practice in high-performance Haskell
programming to manually review GHC's optimized code output.

\subsection{Closure-Free Staged Compilation}

@@

Metaprograms can be deterministic, transparent, and can be run efficiently,
using fast interpreters or machine code compilation. In contrast,
general-purpose optimizers are slower to run, less transparent and less
robust. Also, metaprogramming allows library authors to exploit
domain-specific optimizations, while it is not realistic for general-purpose
optimizers to know about all domains.

On the other hand, metaprogramming requires some additional care and input from
programmers. Historically, there have been problems with ergonomics as well. In
weakly-typed staged systems, code generation might fail \emph{too late} in the
pipeline, producing incomprehensible errors. Or, tooling that works for an
object language (like debugging, profiling, IDEs) may not work for metaprogramming, or
metaprogramming may introduce heavy noise and boilerplate, obscuring the logic
of programs and imposing restrictions on code structure.

@@

We develop a particular two-level type theory for this purpose, which we
call \textbf{CFTT}, short for ``closure-free type theory''. This consists
of
\begin{itemize}
\item A simply-typed object theory with first-order functions, general recursion and
      finitary algebraic data types. This language is easy to optimize and compile
      in the downstream pipeline, but it lacks many convenience features.
\item A standard Martin-Löf type theory for the compile-time language. This
      allows us to recover many features by metaprogramming.
\end{itemize}

Since the object language is first-order, we guarantee that all programs in CFTT
can be ultimately compiled without any dynamic closures, using only calls and
jumps to statically known code. Why focus on closures?  They are the
foundation to almost all abstraction tools in functional programming:
\begin{itemize}
\item Higher-order functions in essentially all functional languages are implemented with closures.
\item Type classes in Haskell use dictionary-passing, which relies on closures for function methods.
\item Functors and first-class modules in OCaml and other ML-s rely on closures.
\end{itemize}
Hence, doing functional programming without closures is a clear demonstration
that we can get rid of abstraction overheads.

Perhaps surprisingly, little practical programming relies essentially on
closures. Most of the time, programmers use higher-order functions for
\emph{abstraction}, such as when mapping over lists, where it is expected that
the mapping function will be inlined. We note though that our setup is
compatible with closures as well, and it can support two separate type formers
for closure-based and non-closure-based (``static'') functions. Having both of
these would be desirable in a practical system. In the current work we focus on
the closure-free case because it is much less known and developed, and it is
interesting to see how far we can go with it.

%% For example, difference lists
%% \cite{TODO} are often implemented with closures, but they can be also
%% implemented as binary trees, with the same programming interface.

%% \todo{Essential closures: CPS? Threaded interpreters?}

%% Whole-program defunctionalization is also possible, and it is notably used by
%% the MLton compiler \cite{TODO}. While this can be practical and effective, it
%% can be also expensive and require whole-program processing. Also, conceptually
%% speaking, it does not eliminate closures but instead makes them more transparent
%% to optimizations. In this paper we do not use defunctionalization.

@@

\begin{itemize}
\item In Section \ref{TODO} we present the two-level type theory CFTT, where the
  object language is first-order simply-typed and the metalanguage is
  dependently typed. The object language admits an operational semantics without
  runtime closures, and can be compiled in a way such that all function calls
  are statically known. We provide a supplementary Agda formalization of the
  operational semantics.
\item In Section \ref{TODO} we build a monad transformer library. We believe
  that this is a good demonstration, because monads and monad transformers are
  the most widely used effect system in Haskell, and at the same time their
  compilation to efficient code can be surprisingly difficult. The main idea is
  to have monads and monad transformers \emph{only} in the metalanguage, and try
  to express as much as possible at the meta level, and funnel the object-meta
  interactions through specific \emph{binding-time improvements}.  We show how
  almost all definitions from the Haskell ecosystem of monad transformers can be
  reused in the meta level of CFTT. We also develop let-insertion, join points
  and case switching on object-level data, in the generality of monad transformers.
\item In Section \ref{TODO} we build a pull-based stream fusion library. Here,
  the motivation is to demonstrate essential usage of dependent types to
  concisely solve a difficult problem, namely guaranteed fusion for unrestricted
  combinations of $\concatMap$ and $\mit{zip}$. We use a state machine
  representation that is based on \emph{sums-of-products} of object-level
  values. We show that CFTT is compatible with a \emph{generativity} axiom,
  which internalizes the fact that metaprograms cannot inspect the structure of
  object-level terms. We use this to show that the universe of sums-of-products
  is closed under $\Sigma$-types. This in turn enables a very
  concise definition of $\concatMap$.
\item
  We adapt the contents of the paper to a Template Haskell, with some
  modifications, simplifications and fewer guarantees about generated code.  In
  particular, Haskell does not have enough dependent types for the simple
  $\concatMap$ definition, but we can still work around this limitation. We also
  provide a more faithful Agda embedding of CFTT and our libraries. Here, the
  object theory is embedded as a collection of postulated operations, and we can
  use Agda's normalization command to print out generated object code.
\end{itemize}

\section{Overview of CFTT}\label{sec:overview-of-cftt}

In the following we give an overview of CFTT features. We focus on informal
explanations and examples here. We first review the meta-level language, then
the object-level one, and finally the staging operations which bridge between
the two.

@@

Formally, $\MTy$ is additionally indexed by universes levels, and we have
$\MTy_i : \MTy_{i+1}$. However, this adds a bit of noise, and it is not very
relevant to the current paper, so we shall omit levels. Note that universe
levels are orthogonal to staging; they are instead used to increase the strength
of the system while maintaining logical consistency.

Throughout this paper we use a mix of Agda and Haskell syntax for
CFTT. Dependent functions and implicit arguments follow Agda. A basic example:
\begin{alignat*}{3}
  &\id : \{\vA : \MTy\} \to \vA \to \vA\\
  &\id = \lam \vx.\, \vx
\end{alignat*}
Here, the type argument is implicit, and it gets inferred when we use the
function. For example, $\id\,\True$ is elaborated to $\id\,\{\Bool\}\,\True$,
where the braces mark an explicit application for the implicit argument.
Inductive types can be introduced using a Haskell-like ADT notation, or with a
GADT-style one:
\begin{alignat*}{3}
  &                                           &&\hspace{4em}\data\,\Bool_\M &&: \MTy\,\where\\
  & \data\,\Bool_\M : \MTy = \True_\M\,|\,\False_\M &&\hspace{4em}\ind\ind \True_\M &&: \Bool_\M\\
  &                                           &&\hspace{4em}\ind\ind \False_\M &&: \Bool_\M
\end{alignat*}
Note that we added an $_\M$ subscript to the type; when analogous types can be
defined both on the meta and object levels, we will sometimes use this subscript
to disambiguate the meta-level version.

We use Haskell-like newtype notation, such as in $\newtype \msf{Wrap}\,\vA =
\msf{Wrap}\,\{\mit{unWrap} : \vA \}$, and also a similar notation for (dependent)
record types, for instance as in
\[\data \msf{Record} = \msf{Record}\,\{\mit{field1} : \vA,\,\mit{field2} : \vB\}.\]

All construction and elimination rules for type formers in $\MTy$ stay within
$\MTy$. For example, induction on meta-level values can only produce meta-level
values.

@@

Since $\VTy$ is a sub-universe of $\Ty$, we have that when $\vA : \VTy$ then also
$\vA : \Ty$. Formally, this is specified as an explicit embedding operation, but
we will use implicit subtyping for convenience.

Second, $\bs{\CTy} : \MTy$ is the universe of \emph{computation types}. This is
also a sub-universe of $\Ty$ with implicit coercions. For now, we only specify
that $\CTy$ contains functions whose domains are value types:
\[ \blank\to\blank : \VTy \to \Ty \to \CTy \]
For instance, if $\Bool : \VTy$ is defined as an object-level ADT, then $\Bool
\to \Bool : \CTy$, hence also $\Bool \to \Bool : \Ty$. However, $(\Bool \to
\Bool) \to \Bool$ is ill-formed, since the domain is not a value type. Let us look at an
example for an object-level program, where we already have natural numbers
declared as $\data \Nat := \Zero\,|\,\Suc\,\Nat$:
\begin{alignat*}{3}
  &\hspace{-1em}\rlap{$\add : \Nat \to \Nat \to \Nat$}\\
  &\hspace{-1em}\add :=\,&& \mathrlap{\letrec \go\,\vn\,\vm := \case \vn\,\of}\\
  &\hspace{-1em}         && \ind \Zero     &&\to \vm;\\
  &\hspace{-1em}         && \ind \Suc\,\vn &&\to \Suc\,(\go\,\vn\,\vm);\\
  &\hspace{-1em}         && \go
\end{alignat*}
Every recursive definition must be introduced with $\letrec$. The general syntax
is $\letrec \vx : \vA := \vt; \vu$, where the $\vA$ type annotation can be
omitted. $\mbf{letrec}$ can be only used to define computations, not values
(hence, only functions can be recursive so far).
